### 2020-06-18 chapters/background.tex
Some Background feedback applied

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine by grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop\footnote{https://hadoop.apache.org} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark\footnote{https://spark.apache.org} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times.

As more and more data became available in the data stream format, the need for stream processing increased. Stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some aggregations are impossible to compute with 100\% accuracy in an efficient manner on unbounded data sets. An example of such is counting distinct elements that flow through the data stream. An immediate solution would be to create a regular set of items or strings and have its length be the number of distinct elements processed. However, this approach is not very efficient and has a linear memory usage growth regarding the number of distinct elements. Therefore, as the number of distinct elements approaches large orders of magnitude we run out of memory.

In order to provide a streaming solution, Apache Spark created the Spark Streaming\footnote{https://spark.apache.org/docs/latest/streaming-programming-guide.html} module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not truly exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be updated in real-time per event being processed. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work. The Venn diagram \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded data set. That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite data set. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.

@@

% boundless aggregations no need for windows
Computations over an unbounded data set are not impossible. For example, computing the \textit{average} of an event's field for all incoming events is computationally feasible. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some logical context into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. Is the average of all temperatures of all machines over an infinite period useful for the detection of overheating equipment? In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for more actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}. By creating such a conceptual range, we realize that in practice we need a bounded data set. Partitioning an infinite data stream to obtain such a finite data set is made through the use of windows.

Gedik et. al claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming a\cite{Wang-Windows-Stream-Processing}ggregations"} \cite{Wang-Windows-Stream-Processing}. More formally, Botan et. al define that \textit{"a window W over a stream S is a finite subset of S"} \cite{Botan-SECRET}.

% 2 dimensions of windows
Windows can be categorized by two dimensions: the window size \textit{W} and the sliding step size \textit{S}, both either tuple or time-based. Tuple-based windows' size is specified by the number of items the window can contain and time-based windows specify the time interval worth of incoming data a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is formally defined as \textit{"the distance between consecutive windows"} \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

There are three types of windows: true sliding, stepping or tumbling windows.
% true sliding: unitary steps
A true sliding window is defined by any given window size \textit{W} and a unitary step (\textit{S}=1), which can be defined as time or tuple-based. At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time. Such a true sliding movement can be observed in Figure \ref{fig:sliding-window} with a tuple-based window of size \textit{W} of three tuples and unitary step.

@@

% stepping windows (per x units over Y window size)
% stepping: step > 1 < window size
A stepping window behaves exactly like a true sliding one but the step size \textit{S} is not unitary. In other words, a stepping window is defined by any window size \textit{W} and steps \textit{S} greater than a unit. Once again, both the window and step sizes, \textit{W} and \textit{S} respectively, can be time or tuple-based. Figure \ref{fig:stepping-window} represents a stepping tuple-based window of size \textit{W} of three and a stepping size \textit{S} of two.

@@

% tumbling: step = window size
A tumbling window, also defined by its window size \textit{W} and step size \textit{S}, is a window where W is equal to S. Hence, tumbling windows partition the stream into non-overlapping chunks with the same size. Since the intersection of different tumbling windows is always empty, each event is said to belong to exactly one tumbling window. An example of a tumbling window can be seen in Figure \ref{fig:tumbling-window} with a tuple-based window of size \textit{W} of three and an equally large step size \textit{S}.

@@

For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size. Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values to a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters. 
Approximate aggregations are required where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time. Thus, approximate aggregators trade precision of results for efficient time and space usage. In Section \ref{sec:pds} we explore approximate aggregators that make such a trade-off.

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: 
\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value. The combine function is here on forward represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.

@@

While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques. In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. Tangwongsan et. al define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. Consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink\footnote{https://flink.apache.org}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:

@@

In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to evict events not in this time interval. The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE.

@@

%\section{Aggregations on data types}

%Some aggregations are only possible on certain data types. For instance, computing an average on a set of strings is not possible. In this Section we present the data types we will work with and the set of possible aggregations to compute on them.

%We consider four data types: \textit{numericals}, \textit{strings}, \textit{dates} and \textit{categoricals}. Numerical variables can be measured. When compared to classical programming data types, for example those of the C programming language, numericals comprise variable types \textit{short}, \textit{long}, \textit{int}, \textit{float}, \textit{double} and valid combinations of the previous --- \textit{e.g.} \textit{long int}. On the other hand, Strings are sequences of characters. The \textit{string} type is the most generic data type in that any other type can be converted to a string. Every variable, for example \textit{x}=3, can be turned into a string \textit{s}, for example \textit{s}="3". Similarly, a date is a sequence of characters and thus a string. However, despite being a string, it has semantics, unlike a regular one. Knowing the date format, one can extract the day, month and year. For that reason we make the distinction between the \textit{string} and \textit{date} data types. Lastly, a \textit{categorical} is a variable that takes one of a relatively small and known set of possible values. A variable is assigned to a particular category based on its properties. The categorical variable is also a sequence of characters and therefore a string variable. The main difference between the categorical data type and the string is that we know the set of possible labels the categorical variable might take and this affects the type of computable aggregations.

%On a set of numerical variables, we can compute sums, minima, maxima, averages, standard deviations and counts (of zeros, infinities and others). When operating on a set of strings, we can aggregate the collection into a concatenation of strings, a count, a membership query, cardinality.

>>>>

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine by grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a distributed computation framework or programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this they read and write from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

On the other hand, stream processing refers to the processing of data streams. A data stream is essentially a continuous or unbounded data set. This data needs to be processed sequentially and incrementally on a event-by-event basis or over sliding windows. Streaming data includes a wide variety of data such as logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices or credit card transactions. This is in contrast to batch processing which works on a finite data set. This distinction is important since some aggregations are very hard and even impossible to compute with 100\% accuracy for several use cases in an efficient manner on unbounded data sets. An example of such is counting distinct elements that flow through the data stream. An immediate solution would be to create a regular set of items or strings and have its length be the number of distinct elements processed. However, this approach is not very efficient and has a linear memory usage growth regarding the number of distinct elements. Therefore, as the number of distinct elements approaches large orders of magnitude we run out of memory.

In order to provide a streaming solution, Apache Spark created the Spark Streaming \cite{SparkStreaming} module. Being a batch system in its core, Spark implements streaming as micro-batching \cite{SparkStreamingPaper}. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not truly exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be updated in real-time per event being processed. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

The Venn diagram \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded data set. That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite data set. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.

@@

% boundless aggregations no need for windows
Computations over an unbounded data set are possible. For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events (one integer) and the current total sum (one double) of all events arriving to the system. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some boundaries into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. The average temperature of a machine over an infinite period (\textit{i.e.} since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure. In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for more actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}. By creating such a conceptual range, we realize that in practice we need a bounded data set. Partitioning an infinite data stream to obtain such a finite data set is made through the use of windows.

\subsection{Streaming Windows}

Botan et. al \cite{Botan-SECRET} formally define a window as:

\begin{definition}
A window W over a stream S is a finite subset of S"
\end{definition}

% 2 dimensions of windows
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is formally defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

Windows can also be characterized by how they move across data, \textit{i.e.}, how the step $S$ is defined. The three most common ways are: true sliding, stepping or tumbling windows. A true sliding window is a window with an unitary step, i.e., where $S = 1$. At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time. Such a true sliding movement can be observed in Figure \ref{fig:sliding-window} with a tuple-based window of size \textit{W} of three tuples and unitary step.

@@

% stepping windows (per x units over Y window size)
% stepping: step > 1 < window size
A stepping window behaves exactly like a true sliding one but the step size \textit{S} is not unitary, \textit{i.e.}, $S > 1$. In other words, a stepping window is defined by any window size \textit{W} and steps \textit{S} greater than a unit. Figure \ref{fig:stepping-window} represents a stepping tuple-based window of size \textit{W} of three and a stepping size \textit{S} of two.

@@

% tumbling: step = window size
A tumbling window is a window with a stepping size as large as the window, \textit{i.e.}, $S = W$. Hence, tumbling windows partition the stream into non-overlapping chunks with the same size. Since the intersection of different tumbling windows is always empty, each event is said to belong to exactly one tumbling window. An example of a tumbling window can be seen in Figure \ref{fig:tumbling-window} with a tuple-based window of size \textit{W} of three and an equally large step size \textit{S}.

@@

For the scope of this Thesis, we will work under a tuple-based true sliding window framework. Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values into a single one. Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality). An aggregation may be exact or approximate. Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists. On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters. 
Approximate aggregations are most useful in situations where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time. Thus, approximate aggregators trade precision of results for efficient time and space usage. In Section \ref{sec:pds} we explore approximate aggregators that make such a trade-off.

Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method \cite{Java13StreamReduce}: 
\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Similarly, the MapReduce \cite{MapReduce} paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one. \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state. Doing this recursively for each item in the collection results in a single value: the aggregate value. The combine function is here on forward represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.

@@

While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded data set from a data stream by applying windowing techniques. When doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et. al} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data cencer in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:

@@

In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element inserted or evicted from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE.

@@

### 2020-06-18 chapters/introduction.tex
Some Background feedback applied

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. Different use cases process data in different ways for different purposes. For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's\footnote{https://alibaba.com} clients, while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. This media content may have been previously recorded and stored, or may be generated on the fly. Social networks are also responsible for generating large volumes of data,Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. These need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the protected system.

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former case is best known as stream processing, while the latter as batch processing.

E-commerce applications have to process endless streams of credit card transactions. Live streaming services need to distribute the media content created by the \textit{streamer} (content creator) and deliver it across multiple users. Social networks process and analyze hundreds of posts per second to find out trending keywords across their network and suggest them to users. Computer network monitor systems want to detect intruders and revoke them system access before they have time to corrupt the system or steal private information. All of the previously given use cases rely on real-time processing of data for near real-time actions. Hence, between batch and stream processing, the latter fits these use cases better.

As mentioned, the processing of data can be divided into two main categories: batch and stream processing. However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing. Data stream processing refers to the handling of huge volumes of high variety of data generated at high-velocity from multiple sources in real-time. Unlike batch processing on static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}. Furthermore, in the streaming paradigm, the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}. \textit{Recency} is measured under different scales for different use cases. For instance, considering the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection. In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe. However, in the intruder detection scenario, the decision of removing access to a user must be done as soon as possible, preferably in a few seconds. Hence, while in the first scenario information retrieved within a day would still be recent, in the second scenario, information is considered recent and relevant if delivered within seconds (or even milliseconds).

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations applied to data sets that produce a single result are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.

Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all incoming transactions and decide in a fraction of a second for each one whether it is fraudulent or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this Thesis, we will work side by side with Feedzai\footnote{https://feedzai.com}. Feedzai offers fraud prevention solutions for the previously presented use case of monitoring an infinite stream of credit card transactions and detecting fraud. We will work with them, taking advantage of the multiple real-life data sets they can provide from the financial fraud space.

>>>>

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. For different use cases, the data is processed in different ways for different purposes. For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. Social networks are also responsible for generating large volumes of data,Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. These need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the protected system.

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter as batch processing.

E-commerce applications have to process endless streams of credit card transactions. Live streaming services need to distribute the media content created by the \textit{streamer} (content creator) and deliver it across multiple users. Social networks process and analyze hundreds of posts per second to find out trending keywords across their network and suggest them to users. Computer network monitoring systems aim to detect intruders and revoke their system access before they have time to corrupt the system or steal private information. All of the referred use cases rely on real-time processing of data for near real-time actions. Hence, stream processing fits these use cases better than batch processing.

As mentioned, the processing of data can be divided into two main categories: batch and stream processing. However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing. Unlike batch processing of static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}. Furthermore, in the streaming paradigm, the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}. \textit{Recency} is measured under different scales for different use cases. For instance, consider the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection. In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe. However, in the intruder detection scenario, the decision of removing access to a user must be done as soon as possible, preferably in a few seconds. Hence, while in the first scenario information retrieved within a day would still be recent, in the second scenario, information is considered recent and relevant if delivered within seconds (or even milliseconds).

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to data sets are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.

Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all incoming transactions and decide in a fraction of a second, for each transaction, whether it is fraudulent or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}. Feedzai offers fraud prevention solutions for the previously presented use case of monitoring a continuous stream of credit card transactions and detecting fraud. We exploit multiple real-life data sets that are provided from the financial fraud space.
